---
title: "R Notebook"
output: html_notebook
---

```[3]
Materials used
S09_multiple.pdf
https://brianreedpowers.github.io/stat340/prediction2.html
```


# Multi Linear Regression

```{r}
data(mtcars)

mtc_model <- lm(qsec ~ 1 + disp + hp + wt, data = mtcars)
summary(mtc_model)

```

```[3]
Estimate: Fitted Coefficients
PT(>|t|): p-value for testing wheather each coffeceient is 0
Residual Standard error: model's typical prediction error size
Multiple R Squared: Proportion of variation explained
F-Statistic: Whether the model as a whole explains anything beyond the intercept only model

For wt, holding disp and hp fixed. A 1 unit increase in wt is associated with an average increase of about 1.49 in qsec

T- value -> estimate / std.error for disp is -1.59. A bigger t-value stronger evidence that the coefficient is not 0.

RSE: The models prediction errors are typically 1.062 seconds
Degrees of Freedom: n is 32, minus 3 predictors + 1 intercept is 28
Multiple R-Squared: About 68% of the variability in qsec is explained by the model using dsip, hp, and wt
F-Statistic: F is 19.91 with a p-value of 4.134 so we reject the null and conclude the model as a whole has signficant explanatory power. And using the variables is much better then using no predictors at all.

We emphasize that the model is usefull overall. But that disp is not useful in the model once hp and wt are already in the model.

qsec = 17.965 - 0.00662(disp) - 0.02295(hp) + 1.485(wt)
```

# 2. Checking assumptions with residuals

```{r}
par(mfrow = c(2,2))
plot(mtc_model)
par(mfrow = c(1,1))
```

```[3]
Residuals vs Fitted (top left)
We are checking linearity and mean of residuals roughly equals 0 with no systemic pattern

The points are mostly scattered aroound 0, the red line has a slight downward pattern with slight curvature
There is no strong violation but suggests possible mild nonlinearity
Acceptable for the class

Q-Q Plot (top right)
Checks if residuals are normally distributed

Middle points lie close to the line
The tails deviate which suggest slight non-normality, especially in the extremes
Normalality assumption isn't perfect but still acceptable for most cases
But be careful with p-values if deviations are too large

Scale Location (bottom left)
Homoscadestacity

Curved uowards, spread increases a bit for higher fitted values

Mild heteroscadasticity, variance slightly increases as fitted values increase
But not severe, still acceptable

Residuals vs Leverage (bottom right)
Influenctial Points

Most points clustered low-average -> normal
A few labeled points (Maserati Bora) -> (~0.5) abd (Merc 230) -> Large residuals
Cook's distance lines shown (dashed)
```

# 3. Categorical predictors with two levels 

```{r}
data(mtcars)

mtcars$am <- factor(mtcars$am) # automatic = 0, manual = 1
mtcars$vs <- factor(mtcars$vs) # v-shaped = 0, straight = 1

cat_model <- lm(qsec ~ 1 + am + vs, data = mtcars)
summary(cat_model)
```

# 4. Interactions between two catagorical predictors

```{r}
## am:vs = interaction only 
## am * vs = main effect plus interactions
library(interactions)


interact_model <- lm(qsec ~ 1 + am * vs, data = mtcars)
summary(interact_model)

cat_plot(interact_model,
         pred = "vs",
         modx = "am",
         geom = "line",
         plot.points = TRUE,
         data = mtcars,
         interval = FALSE)
```
```[3]
Looking at am1:vs1, we can look at the combined effect of being in both categorical values together. 
am1 = manual, and vs1 = staight engine. For cars that are both, the predicted qseec is .07 higher then what you would expect from just adding the seperate main effects of am and vs alone. But it is not significant, p-value .91% so a simpler model qsec ~ am + vs is sufficient
```
# 5. Categorical predictors with more than two levels

```{r}
data(iris)

iris$Species <- factor(iris$Species, levels = c("setosa", "versicolor", "virginica"))

sepal.fit <- lm(Sepal.Length ~ 1 + Species, data = iris)
summary(sepal.fit)

# check the means for each dummy
aggregate(Sepal.Length ~ Species, data = iris, FUN = mean)

# see the dummy vars
model.matrix(sepal.fit)[1:10, ]
```
```[3]
Setosa is the baseline so the intercept is the mean sepal length for setosa

Speciesversicolor = difference between versicolor and setosa
Speciesviginica = difference between virginica and setosa

So 
Setosa: 5.006
Versicolor: 5.006 + 0.96 = 5.936
Virginica: 5.006 + 1.582 = 6.588
```

# 6. Interaction example where interaction matters

```{r}
set.seed(340)

Food <- factor(rep(c("Hot Dog", "Ice Cream"), each = 40))
Condiment <- factor(rep(rep(c("Chocolate", "Mustard"), each = 20), 2))

# Build a strong interaction:
# Hot Dog + Mustard high
# Ice Cream + Chocolate high
Enjoyment <- c(
  rnorm(20, mean = 65, sd = 4),   # Hot Dog + Chocolate
  rnorm(20, mean = 90, sd = 4),   # Hot Dog + Mustard
  rnorm(20, mean = 92, sd = 4),   # Ice Cream + Chocolate
  rnorm(20, mean = 62, sd = 4)    # Ice Cream + Mustard
)

frost_like <- data.frame(Food, Condiment, Enjoyment)

# No interaction
m1 <- lm(Enjoyment ~ 1 + Food + Condiment, data = frost_like)
summary(m1)

# With interaction
m2 <- lm(Enjoyment ~ 1 + Food * Condiment, data = frost_like)
summary(m2)
```

```[3]
What we see in the interaction between Ice Cream and Mustard
In m1, we see that it is misleading because we are averaging between two opposite patterns
In m2, we see that the interaction terms are significant because the condiment effect flips depending on the food

In simple terms we can see that becayse the best condiment depends on the food. Without interaction both main effects can look non-significant; once the interaction is added, everything becomes signficant. 
```

# 7. Nonlinear relationships: adding a squared term

```{r}
plot(mtcars$hp, mtcars$mpg,
     main = "mpg vs hp",
     xlab = "hp", ylab = "mpg")

# Linear model
mtc_lm <- lm(mpg ~ 1 + hp, data = mtcars)
summary(mtc_lm)

par(mfrow = c(1, 2))
plot(fitted(mtc_lm), resid(mtc_lm),
     main = "Residuals vs Fitted: linear",
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, lty = 2)

# Quadratic model
mtc_lm2 <- lm(mpg ~ 1 + hp + I(hp^2), data = mtcars)
summary(mtc_lm2)

plot(fitted(mtc_lm2), resid(mtc_lm2),
     main = "Residuals vs Fitted: quadratic",
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, lty = 2)
par(mfrow = c(1, 1))
```
```{r}
plot(mtcars$hp, mtcars$mpg,
     xlab = "hp", ylab = "mpg",
     main = "Linear vs quadratic fit")

hp_grid <- seq(min(mtcars$hp), max(mtcars$hp), length.out = 200)

lines(hp_grid,
      predict(mtc_lm, newdata = data.frame(hp = hp_grid)),
      lwd = 2)

lines(hp_grid,
      predict(mtc_lm2, newdata = data.frame(hp = hp_grid)),
      lwd = 2, lty = 2)
```

```[3]
Looking at MPG we can see that the relationship is curved. MPG drops quickly at first then levels off. The fix is to adda quadratic term using I(hp^2)
```

# 8. Practice: better predictor in simple regression

```{r}
data(trees)

ht_lm <- lm(Volume ~ 1 + Height, data = trees)
gr_lm <- lm(Volume ~ 1 + Girth, data = trees)

summary(ht_lm)
summary(gr_lm)
```

```{r}
par(mfrow = c(1, 2))

plot(trees$Height, trees$Volume,
     main = "Volume vs Height",
     xlab = "Height", ylab = "Volume")
abline(ht_lm, lwd = 2)

plot(trees$Girth, trees$Volume,
     main = "Volume vs Girth",
     xlab = "Girth", ylab = "Volume")
abline(gr_lm, lwd = 2)

par(mfrow = c(1, 1))
```
# 9. Practice non-linear: geometry intuition

```{r}
gr_sq_lm <- lm(Volume ~ 1 + Girth + I(Girth^2), data = trees)

summary(gr_lm)
summary(gr_sq_lm)

anova(gr_lm, gr_sq_lm)
```
# 10. Clean workflow for and Stat 340 problem

```{r}
### STAT 340 Multiple Regression Walkthrough

### 1) Main multiple regression example
data(mtcars)
mtc_model <- lm(qsec ~ 1 + disp + hp + wt, data = mtcars)
summary(mtc_model)

par(mfrow = c(2, 2))
plot(mtc_model)
par(mfrow = c(1, 1))

rss <- sum(resid(mtc_model)^2)
n <- nrow(mtcars)
p <- 3
rse <- sqrt(rss / (n - (p + 1)))
rss
rse
summary(mtc_model)$r.squared
summary(mtc_model)$adj.r.squared
summary(mtc_model)$fstatistic

### 2) R^2 and correlation
mt.fit <- lm(hp ~ 1 + wt, data = mtcars)
summary(mt.fit)$r.squared
cor(mtcars$wt, mtcars$hp)^2

mt.fit2 <- lm(qsec ~ 1 + disp + wt + hp, data = mtcars)
summary(mt.fit2)$r.squared
cor(mtcars$qsec, predict(mt.fit2))^2

### 3) Categorical predictors and interaction
mtcars$am <- factor(mtcars$am)
mtcars$vs <- factor(mtcars$vs)

cat_model <- lm(qsec ~ 1 + am + vs, data = mtcars)
summary(cat_model)

interact_model <- lm(qsec ~ 1 + am * vs, data = mtcars)
summary(interact_model)

# Optional:
# library(interactions)
# cat_plot(interact_model, pred = "vs", modx = "am",
#          geom = "line", plot.points = TRUE,
#          data = mtcars, interval = FALSE)

### 4) Multiple categorical levels
data(iris)
iris$Species <- factor(iris$Species,
                       levels = c("setosa", "versicolor", "virginica"))

sepal.fit <- lm(Sepal.Length ~ 1 + Species, data = iris)
summary(sepal.fit)
aggregate(Sepal.Length ~ Species, data = iris, FUN = mean)
model.matrix(sepal.fit)[1:10, ]

### 5) Interaction example with simulated food/condiment data
set.seed(340)
Food <- factor(rep(c("Hot Dog", "Ice Cream"), each = 40))
Condiment <- factor(rep(rep(c("Chocolate", "Mustard"), each = 20), 2))
Enjoyment <- c(
  rnorm(20, mean = 65, sd = 4),
  rnorm(20, mean = 90, sd = 4),
  rnorm(20, mean = 92, sd = 4),
  rnorm(20, mean = 62, sd = 4)
)
frost_like <- data.frame(Food, Condiment, Enjoyment)

m1 <- lm(Enjoyment ~ 1 + Food + Condiment, data = frost_like)
m2 <- lm(Enjoyment ~ 1 + Food * Condiment, data = frost_like)
summary(m1)
summary(m2)

### 6) Nonlinear relationship
plot(mtcars$hp, mtcars$mpg, xlab = "hp", ylab = "mpg")

mtc_lm <- lm(mpg ~ 1 + hp, data = mtcars)
mtc_lm2 <- lm(mpg ~ 1 + hp + I(hp^2), data = mtcars)

summary(mtc_lm)
summary(mtc_lm2)

par(mfrow = c(1, 2))
plot(fitted(mtc_lm), resid(mtc_lm),
     main = "Linear residuals", xlab = "Fitted", ylab = "Residuals")
abline(h = 0, lty = 2)

plot(fitted(mtc_lm2), resid(mtc_lm2),
     main = "Quadratic residuals", xlab = "Fitted", ylab = "Residuals")
abline(h = 0, lty = 2)
par(mfrow = c(1, 1))

plot(mtcars$hp, mtcars$mpg, xlab = "hp", ylab = "mpg")
hp_grid <- seq(min(mtcars$hp), max(mtcars$hp), length.out = 200)
lines(hp_grid, predict(mtc_lm, newdata = data.frame(hp = hp_grid)), lwd = 2)
lines(hp_grid, predict(mtc_lm2, newdata = data.frame(hp = hp_grid)), lwd = 2, lty = 2)

### 7) Practice page: trees example
data(trees)
ht_lm <- lm(Volume ~ 1 + Height, data = trees)
gr_lm <- lm(Volume ~ 1 + Girth, data = trees)
summary(ht_lm)
summary(gr_lm)

par(mfrow = c(1, 2))
plot(trees$Height, trees$Volume, main = "Volume vs Height",
     xlab = "Height", ylab = "Volume")
abline(ht_lm, lwd = 2)

plot(trees$Girth, trees$Volume, main = "Volume vs Girth",
     xlab = "Girth", ylab = "Volume")
abline(gr_lm, lwd = 2)
par(mfrow = c(1, 1))

gr_sq_lm <- lm(Volume ~ 1 + Girth + I(Girth^2), data = trees)
summary(gr_sq_lm)
anova(gr_lm, gr_sq_lm)
```


# To do 

```[3]
1. Categorical predictors with more than two levels
2. Interaction examples where interaction matters
3. Non-linear relationships: adding a squared term
4. Practice-Page comparison: better predictor in simple regression
5. Pratice-page nonlinear idea: geometry intuition
6. A clean workflow for any STAT340 regression problem
7. Common exam traps
8. Full Script
9. What each example is teaching us
```


